Numerical Stability of Detonations in White Dwarf Simulations

### 2018-11-23 papers/ignition/paper.tex
Diagnostic quantities
More discussion of self-heating results

The simulation is terminated when the peak temperature on the domain first reaches
$4 \times 10^9$ K, which we call a thermonuclear ignition. This stopping criterion is a
proxy for the beginning of a detonation. Reaching this temperature does not guarantee
that a detonation will begin, and in this study we do not directly address the question
of whether a ignition of this kind always leads to a detonation. Rather, the main question
we investigate here is whether this ignition is numerically converged. A converged ignition
is a prerequisite to having a converged detonation.

@@

\autoref{fig:self-heat-distance} shows our main results.

@@

>>>>

The simulation is terminated when the peak temperature on the domain first reaches
$4 \times 10^9$ K, which we call a thermonuclear ignition. This stopping criterion is a
proxy for the beginning of a detonation. Reaching this temperature does not guarantee
that a detonation will begin, and in this study we do not directly address the question
of whether a ignition of this kind always leads to a detonation. Rather, the main question
we investigate here is whether this ignition is numerically converged. A converged ignition
is a prerequisite to having a converged detonation. We measure two diagnostic quantities:
the time since the beginning of the simulation required to reach this ignition criterion,
and the distance from the contact point of the peak temperature.

@@

\autoref{fig:self-heat-distance} shows our main results. The lowest resolution we consider,
256 km, is typical of the early simulations of white dwarf collisions, and demonstrates a
prompt ignition near the contact point. As the (uniform) resolution increases, the ignition
tends to occur earlier and nearer to the contact point. This trend is not physically meaningful:
all simulations with resolution worse than about 1 km represent the same prompt central ignition,
and as the resolution increases, there are grid points physically closer to the center that can ignite.
However, when the resolution is better than 1 km, the situation changes dramatically: the prompt
central ignition does not occur, but rather the ignition is delayed and occurs further from the contact
point. When we have finally converged, the ignition occurs around 1000 km from the contact point,
about 1 second after contact (contrast to less than 0.05 seconds for the simulation with 1 km resolution).
This story contains two important lessons. First, the required resolution for a converged simulation,
less than 100 m, is out of reach of current 3D supernova simulations. Second, the behavior for
resolutions worse than 1 km qualitatively appears to be converged, and one could perhaps
be misled into thinking that there was no reason to try higher resolutions, which is reason
for caution in interpreting reacting hydrodynamics simulations.

Also in \autoref{fig:self-heat-distance} we show results for a few cases where we add AMR
using the criterion described above. The main takeaway here is that for coarse grid resolutions
worse than 1 km, the addition of AMR yields unreliable results: the coarse grid's approximation
of the physics must simply not be good enough for refinement to do any good. This parallels the
fact that the trend of convergence towards the correct delayed detonation does not begin until
resolutions better than 1 km. However, for a coarse grid resolution of 1 km, AMR does yield
convergence to the correct answer, and at a much cheaper computational cost than the equivalent
uniform grid simulation\marginpar{quantify this}, so AMR does add value if the resolution on the
coarse grid is sufficiently high.

@@

% Add discussion about how none of the conclusions depend on the microphysics
% of how ignitions/detonations actually occur -- this paper is simply about
% self-consistent fluid simulations, i.e. are we even correctly solving
% the equations we wrote down, and not focusing on the harder question
% of whether we're solving the right equations.
